Strong lensing selection effects

### 2023-07-25 paper/seleff.tex
accepted version

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
    {
Strong lenses are a biased subset of the general population of galaxies.
}
  % aims heading (mandatory)
   {
The goal of this work is to quantify how lens galaxies and lensed sources differ from their parent distribution, namely the strong lensing bias.
} 
   % methods heading (mandatory)
   {
We first studied how the strong lensing cross-section varies as a function of lens and source properties. 
Then, we simulated strong lensing surveys with data similar to that expected for Euclid and measured the strong lensing bias in different scenarios.
We focused particularly on two quantities: the stellar population synthesis mismatch parameter, $\asps$, defined as the ratio between the true stellar mass of a galaxy and the stellar mass obtained from photometry, and the central dark matter mass at fixed stellar mass and size.
}
% results heading (mandatory)
   {
Strong lens galaxies are biased towards larger stellar masses, smaller half-mass radii and larger dark matter masses.
The amplitude of the bias depends on the intrinsic scatter in the mass-related parameters of the galaxy population and on the completeness in Einstein radius of the lens sample.
For values of the scatter that are consistent with observed scaling relations and a minimum detectable Einstein radius of $0.5''$, the strong lensing bias in $\asps$ is $10\%$, while that in the central dark matter mass is $5\%$.
The bias has little dependence on the properties of the source population: samples of galaxy-galaxy lenses and galaxy-quasar lenses that probe the same Einstein radius distribution are biased in a very similar way.
}
  % conclusions heading (optional), leave it empty if necessary 
   {
Given current uncertainties, strong lensing observations can be used directly to improve our current knowledge of the inner structure of galaxies, without the need to correct for selection effects.
bf Time-delay measurements of $H_0$ from lensed quasars can take advantage of prior information obtained from galaxy-galaxy lenses with similar Einstein radius.
}
   \keywords{
             Gravitational lensing: strong 
               }

@@

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
In order for a lens-source pair to be included in a strong lensing survey, the following conditions must be met:
\begin{enumerate}
\item The lens must produce multiple images of the source
\item The multiple images must be detected and resolved
\item The system must be recognized as a lens by the survey.
\end{enumerate}
Condition 1 poses constraints on the mass distribution of the lens, as well as on the geometry of the system. 
First of all, not all lenses are capable of producing multiple images, but only those whose surface mass density $\Sigma(\boldsymbol\theta)$ is larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
Second, objects with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged, and therefore are over-represented in lens samples. Third, at fixed lens properties, sources that are located further away along the line of sight are more likely to be lensed.

@@

A measurement of the time delay between the images of a lensed quasar can be converted into an estimate of the expansion rate of the Universe, $H_0$. 
In general, though, the value of $H_0$ is degenerate with the mass distribution of the lens (and of matter along the line-of-sight), which can be difficult to constrain on an individual lens basis.
One possible strategy to break this denegeracy is to use information gathered from a larger set of lenses as a prior for the time-delay lenses.
This was the strategy adopted by \citet{Bir++20} in the most recent measurement of $H_0$ by the TDCOSMO collaboration: they combined a larger sample of extended source lenses with a smaller set of time-delay lenses.
In this section we want to quantify the possible bias on $H_0$ that can result from this approach.

@@

In the time-delay study of \citet{Bir++20}, lenses are described with a mass density profile given by \Eref{eq:mstprofile}.
\citet{Bir++20} defined the profile by first fitting a pure power-law mass model to the strong lensing data, thus constraining the Einstein radius and power-law index $\gamma$, and then applying a mass-sheet transformation of the following kind:
\begin{equation}\label{eq:mst}
\kappa(\boldsymbol\theta) \rightarrow \lmst\kappa(\boldsymbol\theta) + 1 - \lmst.
\end{equation}
The mass-sheet transformation parameter $\lmst$ can be interpreted as a quantity describing a departure from a pure power-law profile.
One possible way of estimating $\lmst$ for our lenses is to fit a power-law to the simulated images and then optimize for the value of $\lmst$ that best matches the true density profile. That, however, would be too time-consuming. Instead, we emulate this process as follows.

>>>>

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
    {
Strong lenses are a biased subset of the general population of galaxies.
}
  % aims heading (mandatory)
   {
The goal of this work is to quantify how lens galaxies and lensed sources differ from their parent distribution, namely the strong lensing bias.
} 
   % methods heading (mandatory)
   {
We first studied how the strong lensing cross-section varies as a function of lens and source properties. 
Then, we simulated strong lensing surveys with data similar to that expected for Euclid and measured the strong lensing bias in different scenarios.
We focused particularly on two quantities: the stellar population synthesis mismatch parameter, $\asps$, defined as the ratio between the true stellar mass of a galaxy and the stellar mass obtained from photometry, and the central dark matter mass at fixed stellar mass and size.
}
% results heading (mandatory)
   {
Strong lens galaxies are biased towards larger stellar masses, smaller half-mass radii and larger dark matter masses.
The amplitude of the bias depends on the intrinsic scatter in the mass-related parameters of the galaxy population and on the completeness in Einstein radius of the lens sample.
For values of the scatter that are consistent with observed scaling relations and a minimum detectable Einstein radius of $0.5''$, the strong lensing bias in $\asps$ is $10\%$, while that in the central dark matter mass is $5\%$.
The bias has little dependence on the properties of the source population: samples of galaxy-galaxy lenses and galaxy-quasar lenses that probe the same Einstein radius distribution are biased in a very similar way.
}
  % conclusions heading (optional), leave it empty if necessary 
   {
Given current uncertainties, strong lensing observations can be used directly to improve our current knowledge of the inner structure of galaxies, without the need to correct for selection effects.
Time-delay measurements of $H_0$ from lensed quasars can take advantage of prior information obtained from galaxy-galaxy lenses with similar Einstein radius.
}
   \keywords{
             Gravitational lensing: strong 
               }

@@

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
In order for a lens-source pair to be included in a strong lensing survey, the following conditions must be met:
\begin{enumerate}
\item The lens must produce multiple images of the source
\item The multiple images must be detected and resolved
\item The system must be recognised as a lens by the survey.
\end{enumerate}
Condition 1 poses constraints on the mass distribution of the lens, as well as on the geometry of the system. 
First of all, not all lenses are capable of producing multiple images, but only those whose surface mass density $\Sigma(\boldsymbol\theta)$ is larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
Second, objects with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged, and therefore are over-represented in lens samples. Third, at fixed lens properties, sources that are located further away along the line of sight are more likely to be lensed.

@@

A measurement of the time delay between the images of a lensed quasar can be converted into an estimate of the expansion rate of the Universe, $H_0$. 
In general, though, the value of $H_0$ is degenerate with the mass distribution of the lens (and of matter along the line-of-sight), which can be difficult to constrain on an individual lens basis.
One possible strategy to break this degeneracy is to use information gathered from a larger set of lenses as a prior for the time-delay lenses.
This was the strategy adopted by \citet{Bir++20} in the most recent measurement of $H_0$ by the TDCOSMO collaboration: they combined a larger sample of extended source lenses with a smaller set of time-delay lenses.
In this section we want to quantify the possible bias on $H_0$ that can result from this approach.

@@

In the time-delay study of \citet{Bir++20}, lenses are described with a mass density profile given by \Eref{eq:mstprofile}.
\citet{Bir++20} defined the profile by first fitting a pure power-law mass model to the strong lensing data, thus constraining the Einstein radius and power-law index $\gamma$, and then applying a mass-sheet transformation of the following kind:
\begin{equation}\label{eq:mst}
\kappa(\boldsymbol\theta) \rightarrow \lmst\kappa(\boldsymbol\theta) + 1 - \lmst.
\end{equation}
The mass-sheet transformation parameter $\lmst$ can be interpreted as a quantity describing a departure from a pure power-law profile.
One possible way of estimating $\lmst$ for our lenses is to fit a power-law to the simulated images and then optimise for the value of $\lmst$ that best matches the true density profile. That, however, would be too time-consuming. Instead, we emulate this process as follows.
